An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-06-23 main.tex
Dagens text

\usepackage{amsthm}
\usepackage{amsmath,mathtools,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

In addition to the costly elimination of the existentially quantified variables,
another problem arises for the common case of solving length constraints in
string solvers. Conjunctions of string constraints lead to the computation of
Parikh images of products of regular languages. Applying the approach
in~\cite{generate-parikh-image} would require the computation of the product
before its Parikh image. In several instances we have observed while solving
real-world string constraints, the computation the product of automata exhausts
the memory of any machine due to the exponential blow-up in size of the product,
quickly becoming intractable as the number of terms in the product increases.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image. This enables us to let both calculations inform each
other, eliminating unnecessary work, and pruning the size of the partial
products considered in the computation for a smaller memory footprint. Moreover,
the method can be used iteratively to tackle smaller chunks of the product
incrementally, thereby avoiding running out of memory. We implement this
calculus, along with the approximate method of~\cite{approximate-parikh}, its
fall-back variant adapted from~\cite{generate-parikh-image}, and an adapter for
the \Nuxmv{} model checker, in a tool that we call \Catra{}. \Catra{} allows
solving constraints over groups of synchronised (products of) Parikh automata,
that is symbolic automata with transitions that increment and decrement integer
registers and that allows checking the constraints on these registers after
executing automata.

@@

\subsection{A demonstration of our approach}

@@

\subsubsection{Example~2: Parikh Automata}

Another application for this generalisation is to express Parikh
automata~\cite{parikh-automata}, and this is indeed what the implementation
described in Section~\ref{sec:implementation} does.

\contents{
 \item what is parikh automata
 \item Give an actual example of a parikh automata
}

@@

\section{Extensions}
\subsection{Backjumping and Learning No-Goods}\label{sec:ext:backjumping}
\subsubsection{An Example}
\subsection{Symbolic Automata}\label{sec:ext:symbolic}

@@

\begin{lstlisting}[caption={An example input file for \Catra{}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), range labels, and their single-character shorthand syntax, register increments, and constraints on the final values of their counters.}, label=lst:input-example]
  /* Declare counters. Initially only integers, but other data-types
 could be considered as well. All counters start at zero. */
counter int x, y, z, length;

// List automata. Automata transitions are labelled with intervals,
// and can specify counter increments/decrements. A counter can only
// be used in at most one automaton.

synchronised {
    automaton A {
    init s0;

    s0 -> s1 [0, 10] { x += 1};

    accepting s1, s2;
    };

    automaton B {
    init t1;

    t1 -> t2 [11, 22] { y += 1};

    accepting t1, t2;
    };
};

automaton C {
    init s0;

    s0 -> s1 [15] { z += 1 };
    s1 -> s1 [15] { z += 1};

    accepting s1;
};

// Specify a constraint on the final values of counters; check whether
// there is a word accepted by all automata for which the final counter
// values satisfy this constraints.
constraint z = 2x + 2y && !(x = 0);
\end{lstlisting}

@@

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances generated by the \Ostrich{} string constraint solver when solving the \Fudge{pyex benchmarks}. \Fudge{The instances are all enormous and have some on average 900 products of 400000 automata with 97 counters each and 9999 constraints}. After generating an initial~\InitialNrBenchmarks{}, we remove~\NrBroken{} misgenerated instances that did not parse, as well as~\NrTrivial{} instances solved in under five seconds by the baseline backend. The benchmarks are run on commit \texttt{\commit}.

@@

\subsection{Threats to Validity}
\contents{
  \item There might be better implementations
  \item We might have over-biased
  \item Our benchmarks might be biased
  \item The underlying automata implementation might be bad and therefore skew
  the results.

>>>>

\usepackage{amsthm}
\usepackage{amsmath,mathtools,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

In addition to the costly elimination of the existentially quantified variables,
another problem arises for the common case of solving length constraints in
string solvers. Conjunctions of string constraints lead to the computation of
Parikh images of products of regular languages. Applying the approach
in~\cite{generate-parikh-image} would require the computation of the product
before its Parikh image. In several instances we have observed while solving
real-world string constraints, the computation of the product of automata
exhausts the memory of any machine due to the exponential blow-up in size of the
product, quickly becoming intractable as the number of automata in the product
increases.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image. This enables us to let both calculations inform each
other, eliminating unnecessary work, and pruning the size of the partial
products considered in the computation for a smaller memory footprint. Moreover,
the method can be used iteratively to tackle smaller chunks of the product
incrementally, thereby avoiding running out of memory. We implement this
calculus, along with the approximate method of~\cite{approximate-parikh}, its
fall-back variant adapted from~\cite{generate-parikh-image}, and an adapter for
the \Nuxmv{} model checker, in a tool that we call \Catra{}. \Catra{} allows
solving constraints over groups of synchronised (products of) Parikh automata,
that is (finite-alphabet) symbolic automata with transitions that increment and
decrement integer registers and that allows checking linear constraints on these
registers after running the automata. Parikh automata without constraints on the
final values on their registers are also sometimes called cost-enriched
automata, weighted automata or counter automata, depending on your definitions
and side-constraints.

\contents{
  \item Automata intersection emptiness is PSPACE-complete
  \item Producing the formula is linear, but
  \item deciding a product of automata on a word is also PSPACE-hard
  \item try proving that finding a word accepted by a product AND satisfying additional constraints is IN PSPACE, e.g. that we are PSPACE-complete
}

@@

\subsection{A demonstration of our approach}\label{sec:introduction:motivation}

@@

\subsubsection{Example~2: Parikh Automata}\label{sec:parikh-automata}

Another application for this generalisation is to express Parikh automata, and
this is indeed what the implementation described in
Section~\ref{sec:implementation} does. Following~\cite{parikh-automata} as
filtered through~\cite{expressiveness} (which offers a better formalisation for
our use), we define a Parikh automaton as an automaton with integer counters
that are incremented at the transitions, and where we have linear
constraints on the final values of the counters after running the automaton, or
formally as in \cref{def:parikh-automata,def:projections}.

In the following definitions, we use (as in~\cite{expressiveness}) the following notation:
\begin{itemize}
  \item $\Alphabet$ is an alphabet
  \item $D$ is a finite subset of $\Naturals^d$, for some $d \in \Naturals^+$ (e.g. vectors of length $d$)
\end{itemize}

\begin{definition}\label{def:projections}
  We define two projections on the composite alphabet $\left(\Alphabet \times D\right)^*$:
  \begin{itemize}
    \item $\pi_{\Alphabet} := (a, \Vector{v}) \mapsto a$ or the projection on $\Alphabet$ is the monoid morphism from $\left(\Alphabet \times D\right)^*$ to $\Alphabet^*$, and the opposite projection
    \item $\pi_{v} := (a, \Vector{v}) \mapsto \Vector{v}$, called the
    \emph{extended Parikh image}.
  \end{itemize}
\end{definition}

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where $C \subseteq \Naturals^d$ is a semi-linear set (or, equivalently, a Presburger formula), where $\Automaton$ is a finite automaton with the alphabet $\Alphabet \times D$, where $D \subseteq \Naturals^d$.

  We say that $\Tuple{\Automaton, \varphi}$ \emph{recognises} the language, $\Language(\Automaton, C)$, that is it recognises the language crated by is the $\pi_\Alphabet$-projection on $\Alphabet$ whose $\pi_{v} \subseteq C$.
\end{definition}

\begin{figure}[ht]
  \centering
      \includegraphics[width=0.75\linewidth]{parikh_automaton}
      \caption{A The automaton part of a Parikh automaton for $\AcaOrBc{}$ with
      $\Alphabet = \Set{\text{a, b, c}}, d = 3$. The semilinear set/Presburger
      formula containing the constraints on the final register values cannot be
      visualised.}\label{fig:parikh-automaton}
    \end{figure}

An example of a Parikh automata version of $\AcaOrBc{}$ can be seen in \cref{fig:parikh-automaton}. A corresponding Presburger version of the constraint seen in \cref{sec:introduction:motivation} is $\psi(\Vec{v}) = v_1 > v_2$.

Generally, we can encode the problem as follows:
\begin{enumerate}
  \item The morphism ($\Map$) we would use is $\pi_{v}$
  \item The monoid operation is element-wise vector addition, which is trivially commutative.
  \item \Fudge{We would just add the constraints on the final values!?}
\end{enumerate}

@@

\section{Extensions}\label{sec:extensions}

This section describes a number of possible extensions to the calculus. Note
that some are partially or fully implemented in \Catra{} already. In particular,
we have support for finite symbolic transitions over Unicode alphabets to keep
practical automata under a reasonable size.

\subsection{Backjumping and Learning No-Goods}\label{sec:ext:backjumping}
\contents{
  \item What do we have
  \item A case where no-goods would be good
  \item What to add to the calculus?
}

\subsubsection{An Example}
\subsection{Symbolic Automata}\label{sec:ext:symbolic}

\contents{
  \item Should we also write about how to adapt to infinite alphabets?
}

\subsection{Transducers}\label{sec:ext:transducers}

\contents{\item What actually goes here!?}

@@

\begin{lstlisting}[caption={An example input file for \Catra{} for the problem introduced in \cref{sec:introduction:motivation}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), labels (except those with ranges), register increments, and constraints on the final values of their counters.}, label=lst:input-example]
  // So far we only support integer counters. Note that we have individual
  // counters for each automaton to avoid surprises for product construction.
  counter int l_a, l_b, l_c, r_c;

  synchronised {
  automaton aca_or_bc {
    init start;

    // We use ASCII values here, this is for lowercase a
    start -> sawA [97] { l_a += 1 };
    start -> sawB [98] { l_b += 1 };

    sawB -> start  [98] { l_b += 1 };
    sawB -> final [99] { l_c += 1 };

    sawA -> sawA [99] { l_c += 1 };
    sawA -> final [97] { l_a += 1 };

    accepting final;
  };

  automaton something_c_something {
      init start;

      // Special short-hand value for the whole Unicode alphabet.
      start -> start [any] ;
      start -> final [99] { r_c += 1 };

      final -> final [any];

      accepting final;
  };
  };

  // Constrain the number of a:s to be larger than the number of c:s. Since the
  // automata are synchronised on every transition and counters are guaranteed
  // to be consistent these constraints are sufficient without also constraining
  // r_c.
  constraint l_a > l_c;
\end{lstlisting}

@@

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances generated by the \Ostrich{} string constraint solver when solving the \Fudge{pyex benchmarks}. \Fudge{The instances are all enormous and have some on average 900 products of 400000 automata with 97 counters each and 9999 constraints}. After generating an initial~\InitialNrBenchmarks{}, we remove~\NrBroken{} misgenerated instances that did not parse, as well as~\NrTrivial{} instances solved in under five seconds by the baseline backend. The benchmarks are run on commit~\texttt{\commit}.

@@

\subsection{Threats to Validity}

The most obvious threats to validity would be poor benchmarking or poor
implementation, e.g. if the method described in \cref{sec:calculus,sec:multiple}
deviate from what is actually benchmarked in section \cref{sec:experiments}, or
if the methods used for benchmarking would be unsound. The first problem cannot
easily be addressed, except by stating that we have been careful in our
implementation and descriptions alike. The second is more practical. To increase
the probability that our results are indeed representative both in the sense of
representing runs on expected, real world input and in the statistical sense for
a given run of \Catra, despite the fact that we use randomness in our
implementation, we execute many experiments. During the development of \Catra{},
we have also executed experimeints on different hardware and software
configurations on smaller samples of instances to ensure that the general trend
holds. The logs from these executions are distributed along with notes as part
of the artefact in the interest of transparency. Moreover, we have validated
\Fudge{all reported satisfiable instances and their witnesses} with \Nuxmv to
ensure that \Calculus{} is indeed sound and does not cheat.

Another threat to validity would be if our implementations of the competition
(baseline and \Nuxmv) would deviate significantly from the expected
implementations. For \Nuxmv{} we use the default configuration which we believe
should be performant (or it should not be the default). Additionally, tweaking our invocation of \Nuxmv is explicitly made easy for artefact reviewers.

The most important and most probable threat to validity is our choice of
implementation platform and our choice of automata library, that is our choice
not to use one. Since our automata are more advanced than anything supported by
e.g. the BRICS library, we opted to use our own implementation. There are some
signs that in particular our product computation is inefficient, notably the
good performance of low-threshold automata materialisation that prioritises
computing smaller products. This means that our implementation of \Calculus{}
makes much less heavy use of our product construction library than the baseline
implementation does, instead putting its weight into \Princess{}' theorem
proving abilities. This situation would unfairly advantage \Calculus{} if our
automata library was indeed the bottleneck. We believe this is unlikely since
similar performance issues as the ones in baseline in our string solver was the
instigating reason for this research. Additionally, while we may have a
constant-factor performance impact from bad implementation, it is less likely
for our potentially poor implementation to be asymptotically worse than e.g.
BRICS. However, the performance improvement of \Calculus{} over baseline seen in
\cref{sec:experiments} is generally not constant.
